\documentclass[./m2r-report.tex]{subfile}

Much of algebra is concerned with substructures of algebraic objects, such as
subgroups, ideals, subspaces, etc.
Almost all of these objects have some inherent structure or 'rules' from which
the existence of these sub-objects can be inferred.
Ramsey theory aims to find conditions for the existence of substructures when
the superstructure has very few rules.
We aim to instil an appreciation of the subject in the reader and show how it may
be applied to other areas of mathematics.

Algebra is primarily concerned with the structure of algebraic objects and their
substructure (subgroups, ideals, subspaces, etc.).
Algebraic objects satisfy a set of axioms (rules) defining their intrinsic
order: existence of ordered substructure can sometimes be inferred by these.
Ramsey theory aims to find conditions for the existence of substructure when the
superstructure satisfies very few rules.

A typical example of an object with little constraint on its structural
properties is a graph.
Graphs are not much more than sets; yet there are many graph theoretical results
of the following form: a large enough graph exhibits regularity.
A notable example is Szemerédi’s regularity lemma, which we state as a theorem:  

\begin{theorem}[Szemerédi]
  For any $\epsilon >0$, any graph admits an $\epsilon$-regular partition.  
\end{theorem}

This theorem is powerful when we consider large, dense graphs\cite{Diestel2000};
$\epsilon$-regularity means in this case that the distribution of edges in our
graph is in some sense homogeneous, and thus “almost random”. This result shows
how a class of seemingly “unstructured” objects can exhibit some strong
structural properties.   

 

Ramsey theory can be seen as a generalisation of this approach to the study of
unstructured or loosely structured mathematical objects. In Ramsey theoretical
terms, a structure exhibits regularity if, no matter how it is partitioned
(coloured), there exists a substructure entirely contained in one class of the
partition. Now, partitioning is a natural operation that can be applied to an
extremely broad class of objects; hence the scope of the theory is very broad.  

 

We aim to instill an appreciation of the subject in the reader and to show how
it may be applied to a wide range of mathematical structures. (can insert the
recap in the abstract here). We will also focus on the ability of the theory to
act as a bridge between combinatorics, algebra and geometry.

% The condition is often a question of quantity, i.e. a substructure will exist
% in every large enough object.


% Local Variables:
% TeX-master: "m2r-report.tex"
% End:

% LocalWords:  Ramsey
