\documentclass[../m2r-report.tex]{subfile}

\subsection{Finite Graphs}
\label{sec:1.1}

Whilst Ramsey's original paper \cite{Ramsey1930} concerned with a problem of logic, Erd{\"o}s and 
Szekeres \cite{Erdos1935} reformulated his work in a graph theoretic sense. The idea of proving 
the existence of regular substructure within an apparently chaotic superstructure
is immediately applicable to graphs. The content of this section primarily follows 
that of Bollobas book \textit{Modern Graph Theory} \cite{Bollobas1998}, and 
I will be using his notation throughout.\\

In the graph below, we partition the edges into two disjoint classes: dotted lines and dashed 
lines. In turns out that no matter how we partition our edges, we will always get
dotted triangles or dashed triangles. In this particular partition, we get two 
dashed triangles.

\begin{figure}[h]
\begin{center}
    \begin{tikzpicture}
        \draw [dotted] (0,0) -- (1,0);
        \draw [dotted] (1,0) -- (1.55,1);
        \draw [dotted] (1.55,1) -- (1,2);
        \draw [dotted] (1,2) -- (0,2);
        \draw [dotted] (0,2) -- (-0.55,1);
        \draw [dotted] (-0.55,1) -- (0,0);
        \draw [dotted] (0,0) -- (1,2);
        \draw [dotted] (1.55,1) -- (-0.55,1);
        \draw [dotted] (1,0) -- (0,2);
        \draw [dashed] (0,0) -- (0,2);
        \draw [dashed] (1,0) -- (1,2);
        \draw [dashed] (1,0) -- (-0.55,1);
        \draw [dashed] (1.55,1) -- (0,2);
        \draw [dashed] (1,2) -- (-0.55,1);
        \draw [dashed] (0,0) -- (1.55,1);

        \filldraw (0,0) circle (2pt);
        \filldraw (1,0) circle (2pt);
        \filldraw (-0.55,1) circle (2pt);
        \filldraw (1.55,1) circle (2pt);
        \filldraw (0,2) circle (2pt);
        \filldraw (1,2) circle (2pt);
    \end{tikzpicture}

\end{center}

\caption{2 dashed triangles within a graph of 6 vertices}
\end{figure}

Before continuing, we first need to state some notation and definitions\cite{Graham1990}.

For $n,k\in \Z^+$, and any non-empty set $X$, let

\begin{align*}%\label{eq:bracketDef}
[n] &= \{1, 2,\ldots, n\},\\
X^{(k)} &= \{Y:Y\subseteq X, |Y|=k\}.
\end{align*}
% At the end of an align, you need to add \\ to go to the next line, I've fixed
% that for you here.
% I'm not sure if you intended to number these equations since I couldn't find
% any references to them in the rest of the section.
% If you did intend to number them, change 'align*' back to 'align' and
% uncomment the label on line 64.
% You can then refer to this definition by \ref{eq:bracketDef} or \hyperlink{eq:bracketDef}{text_to_display}
% QuantumAnon

\begin{definition}[Graph]

    A \textit{graph G} is an ordered pair of disjoint countable sets $(V,E)$ such that $E$ 
    is a subset of $V^{(2)}$, the set of unordered pairs of $V$. We call $V$ the 
    set of vertices, and $E$ the set of edges. The \textit{order}
    of $G$ is the number of vertices, i.e $|V|$.

\end{definition}

\begin{definition}[Complete graph]

    We say a graph $G:=(V,E)$ is \textit{complete} if $E=V^{(2)}$. Intuitively, a graph 
    is complete if every pair of vertices has an edge connecting them. We denoted the complete 
    graph of order $n$ by $K_n$.

\end{definition}

\begin{definition}[Subgraphs and Cliques]

    We say $G':=(V',E')$ is a \textit{subgraph} of $G:=(V,E)$ if $V'\subseteq V$, 
    and $E'\subseteq E$. We write $G'\subseteq G$. We say $G'$ is an \textit{induced} 
    subgraph of $G$ if $E'=E\cap V'^{(2)}$. We say an induced subgraph $G'$ a \textit{clique} of $G$ if it is complete.

\end{definition}

Earlier we mentioned partitioning the edges into disjoint classes. We can think 
of these classes as being `colourings' of the edges. In an \textit{r-colouring}, 
we partition our edges into $r$ disjoint classes, called \textit{colours}. The 
following definition helps explain this:

\begin{definition}[Colouring of a set]

    Let $X$ be a non-empty set. For $r\in\Z^+$, a $r-colouring$ of $X$ is a 
    partition of $X$ into $r$ equivalence classes.
    
    Equivalently, a $r$-colouring of $X$ is the partition of $X$ induced by an 
    arbitrary map: $$\chi: X \to [r]$$ through the equivalence relation: 
    $$\mathcal{R}_{\chi}:= \{(x_{1},x_{2})\in X | \chi(x_{1})=\chi(x_{2}) \}$$
    and $\chi$ is called the colouring function. Somewhat improperly, 
    we will identify a $r$-colouring of $X$ with its colouring function.

    In the context of graphs, an $r$-colouring of the edges of a graph $G=(V,E)$ is a map \cite{Graham1990}:

    $$\chi : E \to [r]$$

    We say a subgraph $G' = (V',E')$ of $G$ is \textit{monochromatic} if the restriction
    $\chi|_{E'}$ is the constant map.

\end{definition}

We can now state the fundamental question that lies at the heart of Ramsey 
theory in a finite graph theoretic sense.

\begin{prop}[Ramsey's theorem for finite graphs]

    Given positive integers $s,t$, there exist a least positive integer $R(s,t)$ 
    such that any 2-colouring of $K_n$ with $n\geq R(s,t)$ contains as a clique 
    either a monochromatic $K_s$ or a monochromatic $K_t$.

\end{prop}

\textbf{Remark:} If $\chi$ is a 2-colouring of the edges of $K_n$, I will refer 
to the edges $\chi^{-1}(\{1\})$ as being `red',
and the edges $\chi^{-1}(\{2\})$ as being `blue'.

\begin{proof}

    We proceed by double induction on $s$ and $t$. Note that $R(s,2) = R(2,s)=s$, 
    since a 2-colouring of $K_s$ either has a blue edge, or 
    every other edge is red. \cite{Bollobas1998}\\

    Hence $R(s,2),R(2,s),R(t,2),R(2,t)$ all exist. We now assume for an induction 
    that $R(s,t-1)$ and $R(s-1,t)$ both exist. Let $n_1:=R(s-1,t), n_2:=R(s,t-1), n := n_1+n_2$, and consider a
    2-colouring of the edges of $K_n$. We need to show that we have
    either a red $K_s$ or a blue $K_t$. So let $x$ be any vertex of
    $K_n$. Since $x$ is joined by edges to $n-1=n_1+n_2-1$ other vertices,
    either there is at least $n_1$ red edges incident with $x$, or there
    are at least $n_2$ blue edges incident with $x$. We will assume that
    the first case holds, since the opposite result follows by symmetry.\\
    
    Consider a clique $K_{n_1}$ joined to $x$ by red edges. If $K_{n_1}$ has a blue $K_t$, then we 
    are done. Otherwise, by the definition of $R(s-1,t)$, the graph
    $K_{n_1}$ contains a red $K_{s-1}$, which forms a red $K_s$ with $x$.

    Hence the result, by induction.

\end{proof}

\textbf{Remark}: 
This result is easily generalised to any finite $r$-colouring of $K_n$:

\begin{theorem}[Ramsey's Theorem for $r$-colours]\label{thr:genRamsey}

    Given $r$ positive integers $s_1, s_2,\ldots, s_r$, there exist a least
    positive integer $R_r(s_1,\ldots,s_r)$ such that any r-colouring of $K_n$ with 
    $n\geq R_r(s_1,\ldots,s_r)$, contains as a clique a monochromatic $K_{s_i}$,
    for some $i\in{1,\ldots r}$, coloured by the $i^{\text{th}}$ colour.

\end{theorem}

\begin{proof}

    We again proceed by induction on $r$. The base case for $R_2(s_1,s_2)$ is
    true by the previous theorem. So we assume by induction that
    $R_{r-1}(s_1,\ldots ,s_{r-1})$ exists. Then in any $r$-colouring of $K_n$ we 
    replace the first two colours by a new colour. If $n$ is sufficiently large,
    then either there is a $K_{s_1}$ coloured with the $i^{\text{th}}$ colour for
    some $i \in \{3,\ldots,r\}$, or else for $m=R(s_1,s_2)$ there is a $K_m$
    coloured with the new colour. I.e. in the original colouring this $K_m$ is 
    coloured with the first two original colours. In the first case we are done,
    and in the second, for $i=1$ or 2, we can find a monochromatic $K_{s_i}$ in
    $K_m$ coloured in $K_m$ coloured with the $i^{\text{th}}$ colour. Hence,
    $$R_r(s_1,\ldots,s_r)\leq R_{r-1}(R(s_1,s_2),s_3,\ldots,s_r)$$

\end{proof}

Whilst we have been discussing Ramsey's theorem in a strictly graph-theoretic sense,
the theory extends beyond that. If we are given a set $V$ (which we 
have been referring to as the set of vertices of a given graph), so far we have been 
looking at partitioning (colouring) the set $V^{(2)}$ (the edges) into $r$ classes. 
However, the theorem still holds when we partition $V^{(k)}$ into $r$ classes, 
for any integer $k$.

\begin{theorem}[Ramsey's Theorem \cite{Ramsey1930}]

    Given $r$ positive integers $s_1, s_2,\ldots, s_r$, there exist a least
    positive integer $R_r(s_1,\ldots,s_r)$ such that any $r$-colouring of $[n]^{(k)}$ with 
    $n\geq R_r(s_1,\ldots,s_r)$, contains as a clique a monochromatic $[s_i]^{(k)}$,
    for some $i\in{1,\ldots r}$, coloured by the $i^{\text{th}}$ colour.

\end{theorem}

\begin{proof}

We omit the proof of this theorem.

\end{proof}

\subsection{Infinite Graphs}
\label{sec:1.2}

The statement of Ramsey's theorem in the introduction, \autoref{thr:Ramsey}
concerns partitioning a countably infinite set $\Gamma$ into $r\in\Z^+$ classes.

\begin{definition}[Infinite graph]

    We define the infinite complete graph, $K_\infty$, as the ordered pair
    $K_\infty:=(\Gamma,\Gamma^{(2)})$.

\end{definition}

\begin{theorem}[Ramsey's theorem for Infinite Graphs\label{thm:ram}]%\cite{lecturenotes}

    For any r-colouring of the edges of $K_\infty$, i.e. 
    $\chi :\Gamma^{(2)}\to\{c_1,\ldots, c_r\}$ there exists 
    within it a countably infinite monochromatic clique. I.e. $\exists
    X \subset K_\infty$ such that $X$ is countably infinite, and $\chi|_{X^{(2)}}$ is the
    constant function.

\end{theorem}

\textbf{Remark}: It should be noted that whilst \autocite{thm:ram} is referring to 
the edges of the infinite graph, $\Gamma^{(2)}$, Ramsey's original theorem states 
that the same result holds for any $r$-colouring of $\Gamma^{(k)}$, for 
any $k,r \in \Z^+$. That is, we can always find a countably infinite monochromatic 
subset $\Delta \subset \Gamma$.

\begin{proof}

    Fix an arbitrary $a_1 \in \Z^+$. Then by the pigeonhole principal, 
    there must exist a countably infinite set $\mathcal{B}_1\subseteq\Z_+\setminus
    \{a_1\}$ such that all of the $a_1-\mathcal{B}_1$ edges (i.e. edges of the
    form $(a_1,b_1)$ with $b_1\in\mathcal{B}_1$ are of the same colour, say $c_1$.\\

    Now again fix an arbitrary $a_2\in\mathcal{B}_1$. Again, we find some countable
    $\mathcal{B}_2\subseteq\mathcal{B}_1$ such that all of the $a_2-\mathcal{B}_2$
    are of the same colour, say $c_2$.\\

    If we repeat this inductively, we obtain a sequence $a_1,a_2,\ldots$ 
    and a sequence of colours $c_1,c_2,\ldots$ such that $\chi(\{a_i,a_j\})=
    c_i$, for $i<j$.\\

    Now again by the pigeonhole principal, since there are finitely many colours,
    there exists an countable subsequence $c_{i_1}, c_{i_2}, \ldots$ that is 
    constant, i.e. each term is equal to $c_i$. Then $\{a_{i_1},a_{i_2}, \ldots\}$ is a
    monochromatic set, coloured by $c_i$.

\end{proof}

This result has the following result from analysis as an interesting corollary:

\textbf{Corollary}: [Bolzano-Weierstrass Theorem]

Let $(x_i)_{i\in \Z_+}$ be a bounded sequence of real numbers. Then it has
a convergent subsequence.

\begin{proof}

    We define a 2-colouring of ${\Z^+}^{(2)}$ by calling $\{i,j\}$ \textit{red} 
    if $x_i <x_j$, and \textit{blue} if $x_i\geq x_j$.

    Then Ramsey's theorem for infinite graphs gives us a countably infinite monochromatic 
    set, which in this case is a monotone subsequence of $(x_i)$. And since it 
    is bounded, by the monotone convergence theorem it must converge.

\end{proof}


\subsection{Graph Ramsey Theory \cite{Bollobas1998}}
\label{sec:1.3}

\begin{definition}[Isomorphism of Graphs]

\end{definition}

\begin{definition}[Generalised Ramsey Number]

\end{definition}



    